---
title: "Lab 1: R Markdown and Data Manipulation"
author: "Hao Wang"
date: Jan.23rd 2017
output: pdf_document
header-includes: \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#setwd("D:/Dropbox/2017 Spring/ 603/lab/Lab 1")
setwd('~/Dropbox/2017 Spring/ 603/lab/Lab 1')
getwd()
```


## Google's R Style Guide:
<https://google.github.io/styleguide/Rguide.xml>


## Rmarkdown
<http://rmarkdown.rstudio.com/lesson-1.html>

R Markdown provides an authoring framework for data science. You can use a single R Markdown file to both

* save and execute code
* generate high quality reports that can be shared with an audience
* Comparing with LaTeX and Sweave, Markdown's grammar is simpler and more firendly for html display.

R Markdown documents are fully reproducible and support dozens of static and dynamic output formats. 

### I recommend using rmardown for coding and taking notes of your analyses. 

Enter your R codes in the blocks. Rmarkdown is also compatible for other languages: Python, SQL, Stan, C++, Java.


For instance
```{r, echo=TRUE, warning=FALSE, message=TRUE}
#Same R language used in blocks. Using # for marking.
#Echo means if you want to show the R codes in the final reports
#warning means if you want to turn the warning message on or off
#message means if you want to report the console result

library(car) #car is a basic package for applied regression
summary(cars) #cars is a default dataset included in the car package
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

- Lab Practice 1: create a .rmd file. Write down your name and set up your working directory. 



## A standard set-up for your codes.
* Date created
* Date lasted modified
* your working directory using setwd('')
* packages

For instance (You should replace with your own working directory)
```{r, echo=TRUE, message=FALSE}
#Hao Wang
#Date created: Jan. 12th / 2017
#Last modified: Jab. 15th
#setwd("D:/Dropbox/2017 Spring/ 603/lab/Lab 1")
#getwd()
options(digits=10)
library(knitr)
library(rmarkdown)
library(foreign)
#library(xlsx)
```

## Read data in R.
R can handle differnt types of data (csv, txt, excel, stata, sas etc.) 


1. Read csv (comma separated values) file: use read.csv function
```{r}
#R can read files directly from the website
mydata.csv <- read.csv('http://www.cyclismo.org/tutorial/R/_static/simple.csv', header = TRUE, sep=",")
write.csv(mydata.csv, "mydata.csv")
#header option means taking the first row as variable names. 
#sep option is determined by the way the data file is separated
```

2. Read txt 

We create a txt data first and then import the txt file to R.
```{r, warning=FALSE}
require(car)
write.table(cars,"cars.txt", sep="\t")
mydata.txt <-read.table("cars.txt")
```


3. Read dta(stata)
   * For stata 12 and earlier (package 'foreign')
   * For stata 13 and later (package'readstata13', this also works for earlier versions)
   
```{r, warning=FALSE}
#install.packages('readstata13')
require(readstata13)
stata14 <-read.dta13('stata14.dta')
stata12 <-read.dta('stata12.dta')
```

4. Read excel

Similarly, we export an excel file and read it.
```{r, warning=FALSE}
#install.packages("xlsx")
library(xlsx)
library(car)
cars.xlsx <- cars
write.xlsx(cars, 'cars.xlsx', sheetName = "Sheet1", row.names =FALSE, col.names = TRUE)
```

- Lab Practice 2: find a data source of either format, load it into R. 


## Dealing with Missing Data
Missing data is very common in our research. Since the missing points are coded differently (for instance: 'NA', '99', '-999'), make sure to check the dataset before analysis.

### Recoding Missing Data
```{r, warning=FALSE}
y <- c(1,2,3,NA)
is.na(y) # returns a vector (FFFT)
#recode other values to missing, speed =10 to NA
mydata <- cars
mydata$speed[mydata$speed ==10] <- NA
mydata$speed
#Code na.omit() listwise delete all the missing observations. Be careful when using this command
#listwisde deletion is BAD! 
#We can code NAs back to 10
mydata$speed[is.na(mydata$speed)] <- 10
mydata$speed
```

##Recode data
### Create new variables
```{r, warning=FALSE}
mydata$speed2 <- (mydata$speed)^2

#This line create dummny variables based on speed
mydata$speed3 <- ifelse(mydata$speed > 10, 
c("slow"), c("quick")) 
mydata$speed3

# another example: create 3 categories 
mydata$speed4[mydata$speed <10] <- "slow"
mydata$speed4[mydata$speed >= 10 & mydata$speed < 18] <- "middle"
mydata$speed4[mydata$speed >= 18] <- "quick"
mydata$speed4
```

##Subsetting data
```{r, warning=FALSE}
require(car)
mydata <- cars
#selecting by observation values
mydata.sub1 <- subset(mydata, subset = speed >= 10)

#selecting by columns
mydata.sub2 <- subset(mydata, select = c(speed))

#reverse selecting
mydata.sub3 <- subset(mydata, select = -c(speed))
```



- Lab Practice 3: In the file you imported to R, code missing values to NA (if already coded as NA, code missing values to -99). Find another variable and code it into a dummy (categorical variable).






