---
title: "3D Plot"
author: "Hao Wang"
date: "January 22, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Data : Boston Housing from MASS package
link: <http://www.clemson.edu/economics/faculty/wilson/R-tutorial/analyzing_data.html>
The Boston data frame has 506 rows and 14 columns.This data frame contains the following columns:
1. crim:
     per capita crime rate by town 
2. zn: 
    proportion of residential land zoned for lots over 25,000 sq.ft. 
3. indus: 
    proportion of non-retail business acres per town 
4. chas: 
    Charles River dummy variable (= 1 if tract bounds river; 0 otherwise) 
5. nox: 
    nitrogen oxides concentration (parts per 10 million) 
6. rm: 
    average number of rooms per dwelling 
7. age: 
    proportion of owner-occupied units built prior to 1940 
8. dis: 
    weighted mean of distances to five Boston employment centres 
9. rad: 
    index of accessibility to radial highways 
10. tax: 
    full-value property-tax rate per $10,000 
11. ptratio: 
    pupil-teacher ratio by town 
12. black:
    1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town 
13. lstat: 
    lower status of the population (percent) 
14. medv:
    median value of owner-occupied homes in $1000 





#3D Scatterplots
You can create a 3D scatterplot with the scatterplot3d package. Use the function scatterplot3d(x, y, z).
```{r, warning=FALSE}
library(MASS)
library(scatterplot3d)
attach(Boston)
scatterplot3d(medv,dis,crim, main="3D Scatterplot")
```


#3D Scatterplot with Coloring and Vertical Drop Lines

```{r}
scatterplot3d(medv,dis,crim,highlight.3d=TRUE,
  type="h", main="3D Scatterplot")
```
 
 
 

# 3D Scatterplot with Coloring and Regression Plane 

```{r}
s3d <-scatterplot3d(medv,dis,crim, pch=16, highlight.3d=TRUE,
       main="3D Scatterplot")
fit <- lm(medv ~ dis + crim) 
s3d$plane3d(fit)
```



#Spinning 3D Scatterplots

You can also create an interactive 3D scatterplot using the plot3d(x, y, z) in the rgl package.. It creates a spinning 3D scatterplot that can be rotated with the mouse. 

```{r}
library(rgl)
plot3d(medv,dis,crim, col="red", size=3)
```

#Surface Plot Using Lattice
You can create a surface plot with the Lattice package using the following codes.

```{r}
library(lattice)
require(MASS)

Boston.df = data.frame(medv = Boston$medv, lstat = Boston$lstat, dis = Boston$dis)
Boston.loess = loess(medv ~ lstat+dis, data = Boston.df,
  degree = 2, span = 0.25)
Boston.fit = expand.grid(list(lstat = seq(1, 40, 1), dis = seq(0, 89, 1)))
medv = predict(Boston.loess, newdata = Boston.fit)

wireframe(medv ~ dis*lstat, data = Boston.fit,
    xlab = "Distance to Working Place", ylab = "Percentage of Low Status Population",
  main = "Surface Boston Housing",
  drape = TRUE,
  colorkey = TRUE,
  screen = list(z = -60, x = -60)
) 
```


 
